\documentclass{article}

\def\npart {IA}
\def\nterm {Lent}
\def\nyear {2024}
\def\nlecturer {Prof. C.\ Warnick}
\def\ncourse {Probability}

\setcounter{tocdepth}{2}

\input{../../../header.tex}
\numberwithin{equation}{section}

\begin{document}
\maketitle{
    \small
    \noindent\textbf{Limits and Convergence}\\
    Sequences and series in $\R$ and $\C$. Sums, products and quotients.
    Absolute convergence; absolute convergence implies convergence. 
    The Bolzano-Weierstrass theorem and applications (the General Principle of Convergence). 
    Comparison and ratio tests, alternating series test.\hspace*{\fill}[6]
    
    \vspace{10pt}
    \noindent\textbf{Continuity}\\
    Continuity of real and complex valued functions defined on subsets of $\R$ and $\C$.
    The intermediate value theorem. A continuous function on a closed bounded interval is bounded and attains its bounds.\hspace*{\fill}[5]

    \vspace{10pt}
    \noindent\textbf{Differentiability}\\
    Differentiability of functions from $\R$ to $\R$. Derivative of sums and products. The chain rule.
    Derivative of the inverse function. Roll's theorem; the mean value theorem. One-dimensional version of the inverse function theorem.
    Taylor's theorem from $\R$ to $\R$; Lagrange's form of the remainder. Complex differentiaion.\hspace*{\fill}[5]
        
    \vspace{10pt}
    \noindent\textbf{Power series}\\
    Complex power series and radius of convergence. Exponential, trigonometric and hyperbolic functions, and relations between them.
    *Direct proof of the differentiability of a power series within its circle of convergence*.\hspace*{\fill}[4]

    \vspace{10pt}
    \noindent\textbf{Integration}\\
    Definition and basic properties of the Riemann integral. A non-integrable function.
    Integrability of monotonic functions. Integrability of piecewise-continuous functions.
    The fundamental theorem of calculus. Differentiation of indefinite integrals.
    Integration by parts. The integral form of the remainder in Taylor's theorem. Impropper integrals.\hspace*{\fill}[6]
}
\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%
% Lecture 1: 19/01/2024
%%%%%%%%%%%%%%%%%%%%%%

\section{Limits and Convergence}
Consider a sequence of real numbers $a_1, a_2, a_3, \cdots$ which we write as $(a_n)$ or $(a_n)_{n=1}^\infty$ for $a_n \in \R$.

\begin{defi}[Limit]
    We say that $a_n \rightarrow a$ as $n \rightarrow \infty$ for some $a \in \R$ if:
    Given $\epsilon > 0$, $\exists N \in \N$ such that $\forall n \geq N$
    \[\abs{a_n - a} < \epsilon\]
\end{defi}

\begin{remark}
    We can freely replace the strict inequality with $\leq$ and it does not change the definition. 
    We can also replace $\epsilon$ with any multiple of $\epsilon$ say $5 \epsilon$.
    We also assume that $\infty \notin \R$, so $a = \infty$ is not allowed.
\end{remark}
\begin{defi}[Increasing and decreasing]\leavevmode
    \begin{enumerate}
        \item If $a_n \leq a_{n+1}, \ \forall n$, then $(a_n)$ is \emph{increasing}
        \item If $a_n \geq a_{n+1}, \ \forall n$, then $(a_n)$ is \emph{decreasing} 
        \item If $a_n < a_{n+1}, \ \forall n$, then $(a_n)$ is \emph{strictly increasing} 
        \item If $a_n > a_{n+1}, \ \forall n$, then $(a_n)$ is \emph{strictly decreasing} 
    \end{enumerate}
\end{defi}

In all cases $(a_n)$ is \emph{monotone}. This allows us to state the \emph{fundamental axiom of real numbers}.

\begin{thm}[Fundamental axiom of real numbers]
    An increasing sequence of real numbers which is bounded above converges (i.e, it has a limit).
    For instance, if $a_n \in \R$ with
    \[a_1 \leq a_2 \leq \cdots \ \text{ and } \ a_n \leq A \ \forall n\]
    There exists $a \in \R$ such that $a_n \rightarrow a$ as $n \rightarrow \infty$.
    Equivalently, a decreasing sequence bounded below has a limit.
\end{thm}

This is equivalent to the \emph{least upper bound axiom}, that states that any non-empty set that is bounded above has a \emph{supremum}.
\begin{defi}[Supremum]
    If $S \subset R$ and $S \neq \emptyset$ we say that
    \[\sup(S) = k\]
    if 
    \begin{enumerate}
        \item $x \leq k \ \forall x \in S$ ($k$ is an upperbound for $S$)
        \item Given $\epsilon > 0 \ \exists x \in S$ such that $x > k - \epsilon$ ($k$ is a least upper bound)
    \end{enumerate}
\end{defi}

A similar definition for greatest lower bound (infimum). 
If $\sup(S) \in S$ we say that $\sup(S)$ is the minimum of $S$ and write
\[\sup(S) = \max(S)\]

\begin{ex}
    Show that the fundamental axiom is equivalent to the least upperbound axiom.
\end{ex}

\begin{nlemma}\leavevmode
    \begin{enumerate}
        \item The limit is unique, i.e; if $a_n \rightarrow a$ and $a_n \rightarrow b$ as $n \rightarrow \infty$ then $a = b$.
        \item If $a_n \rightarrow a$ as $n \rightarrow \infty$ and $n_1 < n_2 < n_3 < \cdots$ is a set of increasing indicies, 
        then $a_{n_j} \rightarrow a$ as $j \rightarrow \infty$. (subsequences converge to the same limit)
        \item If $a_n = c \ \forall n$ then $a_n \rightarrow c$ as $n \rightarrow \infty$
        \item If $a_n \rightarrow a$ and $b_n \rightarrow b$ then $a_n + b_n \rightarrow a + b$ 
        \item If $a_n \rightarrow a$ and $b_n \rightarrow b$ then $a_nb_n \rightarrow ab$
        \item If $a_n \rightarrow a$ and $a_n \neq 0$ and $a \neq 0$ then $\frac{1}{a_n} \rightarrow \frac{1}{a}$
        \item If $a_n \leq A$ and $a_n \rightarrow a$ then $a \leq A$. (note that this is not true if we insist $a < A$)
        \item If $a_n \rightarrow a$ and $c_n \rightarrow a$ as $n \rightarrow \infty$ and
        \[a_n \leq b_n \leq c_n \rightarrow b_n \rightarrow a\]
    \end{enumerate}
\end{nlemma}

\begin{proof}
    We will do (i), (ii) and (v), the others wil be exercises.
    \begin{enumerate}
        \item For any $\epsilon >0$ find $N_1(\epsilon)$ and $N_2(\epsilon)$ such that
        \begin{align*}
            n \geq N_1(\epsilon) \Rightarrow \abs{a_n - a} < \epsilon
            n \geq N_2(\epsilon) \Rightarrow \abs{a_n - b} < \epsilon
        \end{align*}
        Hence if $n \geq \max(N_1, N_2)$ then,
        \begin{align*}
            0 \leq \abs{b - a} &= \abs{b - a_n + a_n - a} \\
            &\leq \abs{b - a_n} + \abs{a_n - a} \\
            &< 2\epsilon
        \end{align*}

        This implies that $\abs{b - a} = 0$, as otherwise we can set $\epsilon = \frac{1}{3} \abs{b - a}$ to find
        \[0 \leq \abs{b - a} < \frac{2}{3} \abs{b - a} \ \contradiction\]

        \item Since $n_j < n_{j + 1} \Rightarrow n_{j+1} \geq n_j + 1$, and hence by induction we have $n_j \geq j$.
        As $a_n \rightarrow a$ as $n \rightarrow \infty$, given $\epsilon >0 \ \exists N = N(\epsilon)$ such that
        \[n \geq N(\epsilon) \Rightarrow \abs{a_n - a} < \epsilon\]
        So if we choose $j \geq N(\epsilon)$ then $n_j \geq j \geq N(\epsilon)$ which implies that
        \[\abs{a_{n_j} - a} < \epsilon\]
        Hence $a_{n_j} \rightarrow a$ as $j \rightarrow \infty$
        
        \item As $a_n \rightarrow a$ and $b_n \rightarrow b$ as $n \rightarrow \infty$ for any $\epsilon > 0 \ \exists N_1 = N_1(\epsilon), N_2 = N_2(\epsilon)$
        such that
        \begin{align*}
            n \geq N_1(\epsilon) \Rightarrow \abs{a_n - a} < \epsilon \\
            n \geq N_2(\epsilon) \Rightarrow \abs{b_n - a} < \epsilon
        \end{align*}
        Now,
        \begin{align*}
            \abs{a_nb_n - ab} &= \abs{a_nb_n - a_nb + a_nb - ab} \\
            &\leq \abs{a_nb_n - a_nb} + \abs{a_nb - ab} \\
            &= \abs{a_n}\abs{b_n - b} + \abs{b}{a_n -a}
        \end{align*}
        If $n \geq N_1$ then $\abs{a_n - a} \leq 1$, therefore
        \[\abs{a_n} = \abs{a_n + a - a} \leq \abs{a_n - a} + \abs{a} \leq (1 + \abs{a})\]
        Thus if $n \geq N_3 = \max(N_1, N_2)$ then,
        \[\abs{a_nb_n - ab} < (1 + \abs{a})\epsilon + \abs{b}\epsilon = (1 + \abs{a} + \abs{b})\epsilon\]
        which can be made arbitrarily small. 
        Therefore
        \[a_nb_n \rightarrow ab\]
    \end{enumerate}
\end{proof}

\end{document}