\documentclass{article}

\def\npart {IB}
\def\nterm {Lent}
\def\nyear {2024}
\def\nlecturer {Prof J. Gog}
\def\ncourse {Variational principles}

\setcounter{tocdepth}{2}

\input{../../../header.tex}

\begin{document}
\maketitle{
    \small
    \noindent Stationary points for functions on $\R^n$. 
    Necessary and sufficient conditions for minima and maxima. Importance of convexity. 
    Variational problems with constraints; method of Lagrange multipliers. 
    The Legendre Transform; need for convexity to ensure invertibility; illustrations from thermodynamics.\hspace*{\fill} [4]

    \vspace{10pt}
    \noindent The idea of a functional and a functional derivative. First variation for functionals, Euler-Lagrange equations, for both ordinary and partial differential equations. 
    Use of Lagrange multipliers and multiplier functions.\hspace*{\fill} [3]

    \vspace{10pt}
    \noindent Fermat’s principle; geodesics; least action principles, Lagrange's and Hamilton's equations for particles and fields. 
    Noether theorems and first integrals, including two forms of Noether's theorem for ordinary differential equations (energy and momentum, for example). 
    Interpretation in terms of conservation laws.\hspace*{\fill} [3]

    \vspace{10pt}
    \noindent Second variation for functionals; associated eigenvalue problem.\hspace*{\fill} [2]
}
\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%
% Lecture 1: 5/10/2023
%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\subsection{Motivation}
\subsection{Preliminaries}
\subsection{Constraints + Lagrange multipliers}

\section{Euler-Lagrange}
\subsection{Euler-Lagrange equations}
\subsection{First integrals of E-L}
\subsection{Fermat's principle}

\section{E-L extensions}
\subsection{E-L with constraints}
\subsection{Several dependant variables}
\subsection{Several independant variables}
\subsection{Higher derivatives}

\section{Convex functions}
\subsection{Convexity}
\subsection{Legendre transform}
\subsection{Application to thermodynamics}

\section{Formalism of dynamics}
\subsection{Least action}
\subsection{Noether's theorem}
\subsection{Hamilton's equations}

\section{The second variation}
\subsection{Second variation}
\subsection{Associated eigenvalue problem}
\subsection{Jacobi condition}

\end{document}