\documentclass{article}

\def\npart {IA}
\def\nterm {Michaelmas}
\def\nyear {2023}
\def\nlecturer {Dr P.\ J.\ O'Donnell}
\def\ncourse {Intro to Mechanics}

\input{../../header.tex}

\begin{document}
\maketitle{
    \begin{center}\begin{minipage}[c]{0.9\textwidth}\centering\footnotesize The lecture list is taken directly from the course schedule. However, it has been shortened to 8 lectures for this session.
    \end{minipage}\end{center}

    \small 
    \noindent\textbf{Lecture 1}\\
    Brief introduction

    \vspace{10pt}
    \noindent\textbf{Lecture 2: Kinematics of a single particle}\\
    Position, velocity, speed, acceleration. Constant acceleration in one-dimension. Projectile motion in two-dimensions.

    \vspace{10pt}
    \noindent\textbf{Lecture 3: Equilibrium of a sinle particle}\\
    The vector nature of forces, addition of forces, examples including gravity, tension in a string, normal reaction (Newton's third law), friction. Conditions for equilibrium.
    
    \vspace{10pt}
    \noindent\textbf{Lecture 4: Equilibrium of a rigid body}\\
    Resultant of several forces, couple, moment of a force. Conditions for equilibrium.
    
    \vspace{10pt}
    \noindent\textbf{Lecture 5: Dynamics of particles}\\
    Newton's second law. Examples of pulleys, motion on an inclined plane.

    \vspace{10pt}
    \noindent\textbf{Lecture 6: Dynamics of particles}\\
    Further examples, including motion of a projectile with air-resistance.

    \vspace{10pt}
    \noindent\textbf{Lecture 7: Energy}\\
    Defintion of energy and work. Kinetic energy, potential energy of a particle in a uniform gravitational field. Conservation of energy.

    \vspace{10pt}
    \noindent\textbf{Lecture 8: Momentum}\\
    Definition of momentum (as a vector), conservation of momentum, collisions, coefficient of restitution, impulse.

    \vspace{10pt}
    \noindent\textbf{Lecture 9: Springs, strings and SHM}\\
    Force exerted by elastic sptings and strings (Hooke's Law). Oscillations of a particle attached to a spring, and of a particle hanging on a string. Simple harmonic motion of a particle for small displacement from equilibrium.

    \vspace{10pt}
    \noindent\textbf{Lecture 10: Motion in a circle}\\
    Derivation of the central acceleration of a particle constrained to move on a circle. Simple pendulum; motion of a particle sliding on a cylinder.}
\tableofcontents


\end{document}